Word Sense Disambiguation\\ \small{Leveraging the NLTK Toolkit and External Tools}

### 2011-04-14 paper/paper.tex
fixing conflicts

\documentclass{article}
\usepackage{url}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{multirow}
\title{Word Sense Disambiguation\\
\small{Leveraging the NLTK Toolkit and External Tools}}
\author{
Alec Story
Ansu Abraham
Craig Frey\\
Dustin Tiedemann
Michael Zhu
Thomas Levine
Whitney Foster\\
}

\newcommand{\naive}{na\"ive}
\newcommand{\Naive}{Na\"ive}

@@

We  chose to implement our system in Python, and chose to use the machine
learning algorithms in NLTK which is also Python-based.  We also used NLTK for
its part of speech tagger, dependency parser, and to import the Senseval data.
We coded the rest of the assignment on our own.

@@

\subsection{Selecting optimal feature combination}
\newcommand\ward{forward} %or maybe backwards

We used a \ward stepwise approach to identifying optimal feature combinations.
We first ran the system with all features on, producing a total of $256=2^8$
combinations.  We then used backwards stepwise regression to identify feature
combinations with a small number of features but with high performance

@@

The NLTK was used to implement two machine learning algorithms, \naive{} Bayes and
decision trees.

@@

The NLTK implements a \naive{} Bayes classifier which chooses the best sense given a
dictionary of features and assigns confidence probabilities to the chosen
sense. \Naive{} Bayes does assume an independence of features. The probability of
each feature contributes to the overall choice.

@@

NLTK provides probability measures for each of its guesses when using the
maximum entropy and the \naive{} Bayes learning methods.  We harnessed these to
determine the cutoff for making a guess, or resorting to a ``U'' if we were not
confident in the tagging.  Occasional unknowns also occur because such unknowns
are present in the training data, and we made no effort to remove them, so the
learning algorithm views them as any other label.

@@

We also used NLTK's part of speech tagger to tag the context, and included those
as we did the literal words in the feature set.  This was trained on NLTK's
\verb+maximum entropy treebank pos tagger+ data, which is included with the NLTK
package.

\subsection{Co-occurrence}

Looks through the training data and compiles a list of most common words for
each item in the data. When the test data is run, the feature looks for the
created vector and uses that instead of trying to make a new one. Next the
extractor takes a window of words around the head word and generates the 0, 1
tuple of features for inclusion to the dictionary. This word list is used as the
keys of the feature vector when using the test data as a list of words that may
co-occur with each item in the data. The value associated with each key in the
feature vector starts out as 0 (for occurring 0 times in the window near a test
word) If a key in the feature vector does occur with a word in the test data,
the value associated with the key changes to 1 even if the key occurs more than
once.  This is necessary because the machine learning algorithm we used does not
attempt to analyze the keys beyond simple equality, so higher numbers than one
could have masked the possibly more important distinction between 0 and not-0.
In other words, keeping track of the instances of occurrence vs clustering them
into 2 groups reduces the sparsity of the data and allows for a higher chance of
accurate classifications based on the limited training data that we have. This
is an especially important consideration since the feature vectors that select
the most common words by looking through all the instances of unique lexical
items in the training data are unsupervised.

@@

\begin{itemize}
\item Classifier (\naive{} Bayes or decision list)
\item Collocation window length (0--8)
\item Co-occurrence handling (on or off)
\item Base word handling (on or off)
\item Bootstrap iterations (0--4)
\item Dependency parsing (on or off)
\end{itemize}

The predictor combination yielding the lowest value of the Akaike information criterion
included the following parameters.

\begin{itemize}
\item Intercept
\item Classifier
\item Collocation window size
\item Co-occurrence
\item Base word handling
\item Classifier$\times$collocation window size interaction
\item Classifier$\times$co-occurrence interaction
\end{itemize}

@@

%The negative interaction terms may indicate that the interacting features provide
similar functionality and blah blah blah

@@

The \naive{} Bayes classifier had better top-range performance.

@@

\subsection{Baseline Results}

We ran two baselines, one with each machine learning algorithm:

\begin{tabular}{l l | r r r}
& grain& precision& recall& attempted\\
\hline
\multirow{3}{*}{Naive Bayes} &
coarse  &  0.737 & 0.379 &   51.39\\
&mixed  &  0.730 & 0.375 &   51.39\\
& fine  &  0.693 & 0.356 &   51.39\\
\multirow{3}{*}{Decision Tree} &
coarse  &  0.612 & 0.612 &  100.00\\
&mixed  &  0.594 & 0.594 &  100.00\\
& fine  &  0.552 & 0.552 &  100.00\\
\end{tabular}

@@

Our run with the highest overall F-measure (overall precision 0.718, overall
recall 0.610), used co-occurrence, collocation out to a distance of 2, base word,
no dependency parsing, and bootstrapped once:

@@

An idea for a feature extractor was to take the most common words in a context
of a lexeme and use those words as features. Based off of the co-occurrence code,
this would generate a vector from the training data, find the most common words
from the test data, and then run the collected data through the classifier.

Unfortunately, this feature extractor did not give very good results. Both the
precision and recall for this feature extractor alone on the EnglishLS test data
were 0.039. This is to be expected since it is essentially the co-occurrence
feature extractor but without grouping data into buckets. The sparsity problem
is even more noticeable in this extractor.

@@

%We're using this now, just on a different corpus, right?
%\subsubsection{Bootstrapping on the Brown Corpus}

%The instances of the target words in the Brown corpus were very few, e.g., activate appeared only 3 times, and bootstrapping didn't seem to be very accurate in general.

@@

Collocation seemed to be fairly powerful at disambiguating, which is to be
expected, since function words near to the target word can be very strong
predictors of how the word is being used in context.  Effectiveness was
strongest with words close to the target, and weaker the farther away we
extended the feature.

>>>>

\documentclass{article}
\usepackage{url}
\usepackage{longtable}
\usepackage{graphicx}
\title{Word Sense Disambiguation\\
\small{Leveraging the NLTK Toolkit and External Tools}}
\author{
Alec Story
Ansu Abraham
Craig Frey\\
Dustin Tiedemann
Michael Zhu
Thomas Levine
Whitney Foster\\
}

@@

We  chose to implement our system in Python, and chose to use the machine
learning algorithms in NLTK which is also Python-based.  We also used
NLTK for its part of speech tagger, dependecy parser, and to import
the Senseval data. We coded the rest of the assignment on our own.

@@

\subsection{Selecting optimal feature combination}
\newcommand\ward{backward} %or maybe backwards

We used a \ward stepwise approach to indentifying optimal feature combinations.
We first ran the system with all features on, producing a total of $256=2^8$
combinations.  We then used backwards stepwise regression to identify feature
combinations with a small number of features but with high performance

@@

The NLTK was used to implement two machine learning algorithms, Naive Bayes and
decision trees.

@@

The NLTK implements a naïve bayes classifier which chooses the best sense given a
dictionary of features and assigns confidence probabilities to the chosen
sense. Naive Bayes does assume an independence of features. The probability of
each feature contributes to the overall choice.

@@

NLTK provides probability measures for each of its guesses when using the
maximum entropy and the naive Bayes learning methods.  We harnessed these to
determine the cutoff for making a guess, or resorting to a ``U'' if we were
unconfident in the tagging.  Occasional unknowns also occur because such
unknowns are present in the training data, and we made no effort to remove
them, so the learning algorithm views them as any other label.

@@

We also used nltk's part of speech tagger to tag the context, and included those
as we did the literal words in the feature set.  This was trained on nltk's
\verb+maximum entropy treebank pos tagger+ data, which is included with the nltk
package.

\subsection{Cooccurrence}

Looks through the training data and compiles a list of most common
words for each item in the data in the form of a vector. When the test
data is run, the feature looks for the previously made vector and uses that instead of
trying to make a new one. Next the extractor takes a window of words
around the head word and generates the 0, 1 tuple of features for
inclusion to the dictionary. This list constitues the features in the
feature vector rather than the list of common words. The value
associated with each key in the feature vector starts out as 0 (for
occurring 0 times in the window near a test word) If a key in the
feature vector does occur with a word in the test data, the value
associated with the key changes to 1 even if the key occurs more than
once.  This is necessary because the machine learning algorithm we
used does not attempt to analyze the keys beyond simple equality, so
higher numbers than one could have masked the possibly more important
distinction between 0 and not-0. In other words, keeping track of the
instances of occurrence vs clustering them into 2 groups reduces the
sparsity of the data and allows for a higher chance of accurate classifications
based on the limited training data that we have. This is an especially
important consideration since the feature vectors that select the most
common words by looking through all the instances of unique lexical
items in the training data are unsupervised.

@@

\begin{itemize}
\item Classifier (naive Bayes or decision list)
\item Collocation window length (0--8)
\item Cooccurrence handling (on or off)
\item Base word handling (on or off)
\item Bootstrap iterations (0--4)
\item Dependency parsing (on or off)
\end{itemize}

The predictor combination yeilding the lowest value of the Akaike information criterion
included the following parameters.

\begin{itemize}
\item Intercept
\item Classifier
\item Collocation window size
\item Coocurrence
\item Base word handling
\item Classifier$\times$collocation window size interaction
\item Classifier$\times$cooccurrence interaction
\end{itemize}

@@

As the regression table above shows, it seems that substituting the
Decision Tree classifier for the Naive Bayes baseline provides the
most significant boost to our f measure. It can also be seen here that
increasing the collocation window will increase the f measure by about 0.003
for each additional word added to the window. Additionally, enabling
the cooccurrence feature extractor with the Naive Bayes classifier
will increase the f measure by a significant amount as well.

The rows that contain negative terms may indicate that the interacting
features provide similar functionality. That is, the enabled features
are interfering with each other so that their final influence on the f
measure may affect the f measure negatively. For instance, the row
showing ``classifiert:collocation'' indicates a run where the Decision
Tree classifier was used and the collocation feature was enabled has
resulted in a lower accuracy. This seems to be the general trend for
all of the interacting features. It may be interesting to note that
the interaction between collocation and cooccurrence1 as well as
collocation and dependency parsing produce the most innocuous
cumulative effect. Perhaps this may be why collocation and
cooccurrence are often advocated to be used together.

@@

more discussin of graphs

@@

@@

@@

An idea for a feature extractor was to take the most common words in a context
of a lexeme and use those words as features. Based off of the cooccurrence code,
this would generate a vector from the training data, find the most common words
from the test data, and then run the collected data through the classifier.

Unfortunately, this feature extractor did not give very good results. Both the
precision and recall for this feature extractor alone on the EnglishLS test data
were 0.039. This is to be expected since it is essentially the cooccurrence
feature extractor but without grouping data into buckets. The sparsity problem
is even more noticeable in this extractor.

@@

%We're using this now, just on a different corpus, right?
%\subsubsection{Bootstrapping on the Brown Corpus} 
%The instances of the target words in the Brown corpus were very few, e.g., activate appeared only 3 times, and bootstrapping didn't seem to be very accurate in general.

@@

Collocation seemed to be a fairly powerful disambugator, which is to be
expected, since function words near to the target word can be very strong
predictors of how the word is being used in context.  Effectiveness was
strongest with words close to the target, and weaker the farther away we
extended the feature.
